A Detailed Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2023-10-25 paper/paper.tex
Updated writing.

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}

@@

% ############################################################
% Pre-Processing
% ############################################################
% Look at the benchmarked tool and the information needed to abstract those tools.
% Discussion of the pre-processing steps.
% Explain their computation and I/O requirements.
\subsection{sMRI Pre-Processing}
\subsubsection{Intensity correction}
% NU4 bias correction (Check ANTS brainExtraction)

@@

\subsubsection{Skull-Stripping}
% ANTS brainExtraction

\subsubsection{Spatial Normalization}
% ANTS registrationSyN

@@

\subsection{dMRI Pre-Processing}
\subsubsection{Eddy Currents Distortion Estimation}
% FSL eddy

@@

\section{Results}
% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspots)
% Any key finding: common point between functions, mem bound, fops, etc.

@@

\subsection{Makespan}
% Identify which tools take the longest to execute.
% This will let us decided which pipeline to investigate more.
Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
% TODO Stat test between mean for single and double precision.
\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Profiling of the pipelines using 1 thread.}
	\label{fig:makespan-1thread}
\end{figure}

@@

% TODO FreeSurfer 1-thread sub-figure

In figure~\ref{subfig:hotspots-32threads-freesurfer-reconall}, we observe that OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. The only function shown from FreeSurfer recon-all is the same as the first function from figure~\ref{subfig:hotspots-1thread-freesurfer-reconall}. However, it only accounts for 2\% of the total makespan compared to 3.8\% while running sequentially.

\begin{figure*}[ht!]
	\centering
	% ANTS brainExtraction
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-brainExtraction}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-brainExtraction-fp}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
	\end{subfigure}

	% ANTS registrationSyN
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-registrationSyN}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-registrationSyN-fp}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
	\end{subfigure}

	% FreeSurfer recon-all
	\begin{subfigure}[t]{\textwidth}
		\caption{}
		\label{subfig:hotspots-1thread-freesurfer-reconall}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
	\end{subfigure}

	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-32threads-freesurfer-reconall}
		\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
	\end{subfigure}
	\hfill
	% FSL FAST
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-fsl-fast}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-fast.png}
	\end{subfigure}

@@

% FreeSurfer recon-all
\csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
\csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}

>>>>

\usepackage{cite}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{graphicx}
\graphicspath{{figures}}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{singlelinecheck=false, skip=0pt}
\usepackage{listings}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage[binary-units=true]{siunitx}
\usepackage{csvsimple}
\usepackage{booktabs}

@@

% ############################################################
% Pre-Processing
% ############################################################
% Look at the benchmarked tool and the information needed to abstract those tools.
% Discussion of the pre-processing steps.
% Explain their computation and I/O requirements.
\subsection{sMRI Pre-Processing}
\subsubsection{Intensity correction}
% N4 bias correction (Check ANTS brainExtraction)
% https://ieeexplore.ieee.org/document/5445030

@@

\subsubsection{Spatial Normalization}
% ANTS registrationSyN
% SyN Registration: https://pubmed.ncbi.nlm.nih.gov/17659998/

\subsubsection{Skull-Stripping}
% ANTS brainExtraction
% 1. Truncate image intensity
% 2. N4 bias correction
% 3. SyN Registration
% 4. Warp prior to image using affine
% 5. Atropos: https://pubmed.ncbi.nlm.nih.gov/21373993/

@@

\subsection{dMRI Pre-Processing}
% TODO: Decide which pipelines to use
% \subsubsection{Eddy Currents Distortion Estimation}
% % FSL eddy

@@

\section{Results}
\subsection{Makespan}
Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
We observe significant computation requirements between the pipelines.
\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Profiling of the pipelines using 1 thread.}
	\label{fig:makespan-1thread}
\end{figure}

\subsection{Long-Tail Distribution}
Figure~\ref{fig:longtail} depicts the distribution of average CPU time for each function across all pipelines. We observe a long tail behavior for the execution time. 
% TODO add statistics about the long tail distribution
This confirms that only a small fraction of the functions contribute to a majority of the execution time of MRI preprocessing pipelines.

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{figures/global_longtail.png}
	\caption{Distribution of CPU Time for each function across all pipelines.}
	\label{fig:longtail}
\end{figure}

\subsection{Linear Interpolation}
% TODO
% - Hotspot analysis for ANTS brainExtraction, ANTS registerSyN, and FreeSurfer recon-all
% Interpolation plays major role for volumetric. Mention top function in ANTS and FreeSurfer.
% - What is the cumulative contribution of interpolation in both applications?
% -

\subsection{FreeSurfer parallelism}
% TODO
% - Add results from dummy OMP dynamic scheduling.
% Might want to add those to the discussion instead.
Our results showed poor parallelism while using FreeSurfer with 32 threads. Therefore, we expanded our methods to execute FreeSurfer recon-all with 1, 2, 4, 8, 16, and 32 threads. This allow us to study the multi-threading performance of the application.

Figure~\ref{fig:freesurfer-parallelism} shows the makespan (blue) and parallelism efficiency (red) while varying the allocated number of threads for FreeSurfer. We set the number of threads using the FreeSurfer ``-threads'' flag and the ``OMP\_NUM\_THREADS'' environment variable. While the makespan reduce with the number of threads, the application shows poor parallelism performance. Table~\ref{tab:freesurfer-parallelism} shows decreasing parallelism efficiency from 56.16 to 4.93, for 2 to 32 threads respectively.
\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-freesurfer.png}
	\caption{FreeSurfer: Makespan (Blue) and parallelism efficiency (Red) while varying the number of threads.}
	\label{fig:freesurfer-parallelism}
\end{figure}
\begin{table}[h]
	\centering
	\begin{tabular}{@{}lcccccc@{}}
		\toprule
		\# of Threads       & \multicolumn{1}{l}{1} & \multicolumn{1}{l}{2} & \multicolumn{1}{l}{4} & \multicolumn{1}{l}{8} & \multicolumn{1}{l}{16} & \multicolumn{1}{l}{32} \\ \midrule
		Speedup             & -                     & 1.12                  & 1.33                  & 1.47                  & 1.58                   & 1.58                   \\
		Parallel Efficiency & -                     & 56.16                 & 33.17                 & 18.33                 & 9.88                   & 4.93                   \\ \bottomrule
	\end{tabular}
	\caption{FreeSurfer: Makespan and parallelism efficiency while varying the number of threads}
	\label{tab:freesurfer-parallelism}
\end{table}

The hotspot analysis for FreeSurfer recon-all with 1 and 32 threads are shown in Figure~\ref{fig:hotspots-freesurfer-reconall}. We observe a clear difference in the profiling between the sequential and multi-threaded executions of the pipeline. On the one hand, figure~\ref{subfig:hotspots-1thread-freesurfer-reconall} shows that the sequential execution has dozens of functions and modules contributing to 80\% application execution time. On the other hand, figure~\ref{subfig:hotspots-32threads-freesurfer-reconall} shows that the multi-threaded execution only has 2 modules and 3 functions contributing a majority of the application runtime.

The primary bottleneck in multi-threaded FreeSurfer is the OMP library, which awaits thread synchronization throughout the pipeline. This accounts for at least 76\% of the application runtime. We think this is due to imbalance in the task scheduling and suppose that optimizing the OMP scheduling could significantly benefit FreeSurfer.
\begin{figure}[ht!]
	\centering
	% FreeSurfer recon-all
	\begin{subfigure}[t]{\linewidth}
		\caption{}
		\label{subfig:hotspots-1thread-freesurfer-reconall}
		\includegraphics[width=\linewidth]{figures/hotspots-1thread-freesurfer-reconall.png}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\caption{}
		\label{subfig:hotspots-32threads-freesurfer-reconall}
		\includegraphics[width=\linewidth]{figures/hotspots-32threads-freesurfer-reconall.png}
	\end{subfigure}
	\caption{FreeSurfer recon-all: Hotspot analysis for 1 and 32 threads.}
	\label{fig:hotspots-freesurfer-reconall}
\end{figure}

FreeSurfer plays an integral role in the fMRIPrep workflow. Therefore, the OMP bottleneck shown previously are also found in multi-threaded fMRIPrep, as Figure~\ref{fig:hotspots-fmriprep-32threads} depicts.
\begin{figure}[ht!]
	\centering
	\includegraphics[width=\linewidth]{figures/hotspots-32threads-fmriprep-full.png}
	\caption{fMRIPrep (32 threads): Hotspot analysis.}
	\label{fig:hotspots-fmriprep-32threads}
\end{figure}

\subsection{ANTS single vs. double precision}
% TODO
% - Create makespan plot with only ANTS pipeline
% There's no need to show the hotspot analysis as there are no differences between single and double precision.
% - Results from toy example with single/double precision in ITK
\begin{figure*}[ht!]
	\centering
	% ANTS brainExtraction
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-brainExtraction}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-brainExtraction-fp}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
	\end{subfigure}
	% ANTS registrationSyN
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-registrationSyN}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-ants-registrationSyN-fp}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
	\end{subfigure}
	\caption{ANTS: Single vs. Double precision}
	\label{fig:ants-fp-comparison}
\end{figure*}

\section{old Results}
% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspots)
% Any key finding: common point between functions, mem bound, fops, etc.

@@

@@

% TODO FreeSurfer 1-thread sub-figure
% Note:
% - Way more functions
% - Top 5 accounts for only 15% of total time. Much less than other applications.
% - Way more memory bound that previous application.
% - Focusing on Top 5 and memory bound could give good performance boost.

\begin{figure*}[ht!]
	\centering
	% % ANTS brainExtraction
	% \begin{subfigure}[t]{0.49\textwidth}
	% 	\caption{}
	% 	\label{subfig:hotspots-ants-brainExtraction}
	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction.png}
	% \end{subfigure}
	% \hfill
	% \begin{subfigure}[t]{0.49\textwidth}
	% 	\caption{}
	% 	\label{subfig:hotspots-ants-brainExtraction-fp}
	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-brainExtraction-fp.png}
	% \end{subfigure}

	% % ANTS registrationSyN
	% \begin{subfigure}[t]{0.49\textwidth}
	% 	\caption{}
	% 	\label{subfig:hotspots-ants-registrationSyN}
	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN.png}
	% \end{subfigure}
	% \hfill
	% \begin{subfigure}[t]{0.49\textwidth}
	% 	\caption{}
	% 	\label{subfig:hotspots-ants-registrationSyN-fp}
	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-ants-registrationSyN-fp.png}
	% \end{subfigure}

	% % FreeSurfer recon-all
	% \begin{subfigure}[t]{\textwidth}
	% 	\caption{}
	% 	\label{subfig:hotspots-1thread-freesurfer-reconall}
	% 	\includegraphics[width=\textwidth]{figures/hotspots-1thread-freesurfer-reconall.png}
	% \end{subfigure}

	% \begin{subfigure}[t]{0.49\textwidth}
	% 	\caption{}
	% 	\label{subfig:hotspots-32threads-freesurfer-reconall}
	% 	\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall.png}
	% \end{subfigure}
	\hfill
	% FSL FAST
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-fsl-fast}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-fast.png}
	\end{subfigure}

@@

% FreeSurfer recon-all
% \csvtable{tables/hotspots-1thread-freesurfer-reconall.csv}{FreeSurfer recon-all (1 thread): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-1thread-freesurfer-reconall}
% \csvtable{tables/hotspots-32threads-freesurfer-reconall.csv}{FreeSurfer recon-all (32 threads): Top functions accouting for 80\% of the application makespan.}{extab:hotspots-32threads-freesurfer-reconall}
